The effect of habitat choice on evolutionary rescue in subdivided populations

### 2020-06-13 ms/main.tex
added SI and more text edits

\documentclass[a4paper,11pt]{article}

\usepackage[utf8x]{inputenc}	
\usepackage{fullpage}		
\usepackage[english]{babel}			
%\usepackage{ulem}
\usepackage{xcolor}						
\usepackage{amsmath,amstext,amssymb,amsfonts,amsthm,mathrsfs,chemarrow,stackrel,dsfont}   % Math packages
\usepackage{graphicx} 					
\usepackage{subcaption}					
%\usepackage{cite}	
\usepackage[authoryear,sectionbib,sort]{natbib} % Bio citation format
\linespread{1.7}
\usepackage{titlesec}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}

@@

\usepackage[pdftex, breaklinks, colorlinks, citecolor=black, urlcolor=black, linkcolor=black]{hyperref} 
\usepackage{fourier} % FD I try to avoid the default LaTeX font for non math journals
\usepackage{FiraSans}
\usepackage[T1]{fontenc}

@@

%opening
\title{The effect of habitat choice on evolutionary rescue in subdivided populations}
\author{Peter Czuppon$^{1,2,\ast}$, Fran\c{c}ois Blanquart$^{1,3}$, Hildegard Uecker$^{4,\dag}$,\\ Florence D\'{e}barre$^{2,\dag}$}
\date{}

@@

@@

\linenumbers{}
\modulolinenumbers[2]

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%% INTRODUCTION %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section*{Introduction}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Evol rescue and dispersal %%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In mathematical models, evolutionary rescue is often studied in a spatially homogeneous situation where the whole population experiences a sudden decrease in habitat quality. In this setting, a large number of theoretical results have been established, for example on the effects of recombination \citep{uecker_2015} and horizontal gene transfer \citep{tazzyman_2014}, reproduction mechanisms \citep{glemin_2013,uecker_2017}, intra- and interspecific competition \citep{osmond_2013}, predation pressure \linelabel{R1-5}\chg{\citep{yamamichi_2015}}, bottlenecks \citep{martin_2013}, different genetic pathways \citep{osmond_2019}, and the context-dependent fitness effects of mutations \citep{anciaux_2018}. \chg{In contrast to these abrupt change scenarios, evolutionary rescue can also be studied in a gradually changing environment \citep[e.g.][]{osmond_2017}}\footnote{\pete{Does anybody know other references?} \hildegard{What about Lynch et al. 1991. Adaptive and  demographic responses of plankton populations to environmental change. Limnology and Oceanography 36:1301--1312., B{\"u}rger and Lynch 1995. Evolution and extinction in a changing environment: a quantitative genetic analysis. Evolution 49:151--163 and Lande and Shannon 1996. The role of  genetic  variation  in adaptation and population persistence in a changing environment. Evolution 50:434--437? But please check; it's a long time that I read those...}}. 
Such gradual changes can in particular occur in fragmented environments.

@@

The population living in this environment consists of asexually reproducing, haploid individuals; generations are discrete and non-overlapping. There are two possible types of individuals, wild types and mutants. The individuals go through the following life-cycle: \begin{enumerate}[label = (\roman*)]
	\item Dispersal: individuals may move between patches. Further details about this step are given below.
	\item Reproduction: individuals reproduce within patches. \linelabel{R2-7}The number of offspring produced by an individual of type $i$ in habitat $k$ (before density regulation, if any), i.e.\ its fecundity, is drawn in a Poisson distribution of expectation $\omega_i^k$. Having fewer than $1$ offspring in expectation means that the local subpopulation will get extinct in the absence of immigration, because the deaths of the parents at each generation are not compensated by enough births on average. On the contrary, the local population is viable if the expected fecundity is greater than $1$. We assume that wild-type and mutant populations are viable in old-habitat patches, and that the mutant's expected fecundity there is lower than the wild type's: $1 \leq \omega^{\text{old}}_m < \omega^{\text{old}}_w$ \footnote{\florence{In the table, it is $0 \leq $, yet the populations are supposed to be viable}}. In new-habitat patches, a wild-type population will eventually go extinct, while a mutant one would persist, hence the term ``rescue mutant'': $\omega^{\text{new}}_w < 1 < \omega^{\text{new}}_m$. 
	All parents die at the end of this step.
	\item \linelabel{R2-9} Mutation: wild-type offspring mutate to the rescue mutant type with probability $\theta$ (back mutations from the mutant to the wild type are neglected).
	\item \linelabel{R2-8}Regulation: if the number of offspring produced locally exceeds the local carrying capacity $K_k$ (where $k$ refers to the habitat type, old or new), the population size is down-regulated to the $K_k$: individuals are randomly removed until the local population size is equal to $K_k$. One mutant offspring has the same chance of being removed as one wild-type offspring: \linelabel{AE-3} we assume that wild-type and mutant individuals are competitively equivalent. If the number of offspring is below the carrying capacity, the regulation step is ignored. 
	At the end of this step, all offspring become adults, and a new cycle then begins.
\end{enumerate}

@@

\subsection*{Analysis steps}
We decompose our analysis into several steps of increasing complexity. 
\begin{enumerate}
	\item We first consider an environment that is constant over time, and heterogeneous over space, with a fraction $f_{\text{old}}$ of old-habitat patches and $1-f_{\text{old}}$ of new-habitat patches. The population is initiated with wild-type individuals at carrying capacity in old-habitat patches, and at the migration-selection equilibrium $\widehat{N}_w^{\text{new}}$ in new-habitat patches (see Section S1 in the SI for details), 
	and with a single mutant individual, either in an old- or in a new-habitat patch. There are no further mutations ($\theta = 0$), and we compute the probability of establishment of a mutant lineage. 
	\item We then consider the same environmental setting, but initialize the population with only wild-type individuals. Mutants can appear by mutation during the simulation ($\theta >0$). We compute the probability of adaptation, i.e. that, during a fixed time interval, a mutant appears by mutation and then establishes. 
	\item Finally, we consider the full scenario where each patch degrades one after the other, as described above. The environment is spatially and temporally variable. The population is initialized with only old-habitat patches, all at carrying capacity, with wild-type individuals only. We compute the probability of rescue, i.e. that a mutant appears by mutation and establishes before the population goes extinct. 
\end{enumerate}

@@

Our calculations assume that the subpopulations in old-habitat patches are always at carrying capacity, i.e.\ that there are always enough offspring that are produced to at least replace all the parents. We also consider that mutants are rare compared to wild-type individuals in old-habitat patches. Temporarily ignoring mutations, and denoting by $\widetilde{N}_{i}^{k}$ the number of type-$i$ individuals in a $k$-habitat patch right after dispersal, then the expected number of mutants in this old-habitat patch at the beginning of the next generation is 
%
\begin{equation}\label{eq:sold}
K_{\text{old}}\, \frac{\omega^\text{old}_m \widetilde{N}^{\text{old}}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w + \omega^\text{old}_m \widetilde{N}^{\text{old}}_m}
%
\approx K_{\text{old}}\, \frac{\omega^\text{old}_m \widetilde{N}^{\text{old}}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w} 
%
= \left(1 + a_{\text{old}}\right) \widetilde{N}^{\text{old}}_m \, .
\end{equation}
%
The approximation results from the assumption that mutants are rare compared to wild-type individuals in old-habitat patches. Eq.~\eqref{eq:sold} defines the per-capita expected growth rate of mutants in old-habitat patches, $a_{\text{old}}$. \hl{The sign of this local growth rate depends on the value of $\widetilde{N}^{\text{old}}_w$, the size of the local wild-type population right after dispersal. Mutants are less fecund that wild-type individuals in old-habitat patches ($\omega^\text{old}_m < \omega^\text{old}_w$), yet the number of mutants can increase in old-habitat patches if $\widetilde{N}^{\text{old}}_w$ is low enough. In this case indeed, the population is far from carrying capacity, and the offspring of all types of individuals can make it to the next generation. Competition is fiercer as carrying capacity is nearer. }\footnote{\florence{I think this is more Results than Model}}. \linelabel{R1-14}\chg{The population dynamics of the wild type are analyzed explicitly in Section~1 of the Supplementary Information (SI).}

In new-habitat patches, on the contrary, we assume that population size remains much lower than carrying capacity, so that all offspring make it to the next generation (no regulation is necessary after reproduction).\footnote{\florence{I am getting rid of notation $r$}} The expected number of mutants in a new-habitat patch at the beginning of the next generation, given that there are $\widetilde{N}^{\text{new}}_m$ mutants in the patch right after dispersal, is 
%
\begin{equation}\label{eq:defanew}
\omega_{m}^{\text{new}} \widetilde{N}^{\text{new}}_m = \left(1+a_{\text{new}} \right) \widetilde{N}^{\text{new}}_m,
\end{equation}
which defines $a_{\text{new}}$, the per-capita growth rate of mutants in new-habitat patches. This assumption constrains the differences in carrying capacities among old- and new-habitat patches under which our analysis remains valid. In particular, $K_{\text{new}}$ cannot be too much lower than $K_{\text{old}}$, because in this case, the flow of migrants from old- to new-habitat patches would substantially increase local population sizes in new-habitat patches.

\linelabel{R1-17}\linelabel{R2-13} These assumptions are relaxed in our stochastic simulations.

@@

We investigate the effect of the different dispersal schemes on the probability of successful adaptation and evolutionary rescue. We first compute the \textit{establishment probability} of a single mutant individual arising either in an old- or in a new-habitat patch. We then link this probability to the dynamics of a source-sink system, i.e. a fixed environment with a certain number of old habitats (sources) and new-habitat patches (sinks). We derive an expression for the probability of a mutation to emerge and establish in a given time interval. We call this second quantity the \textit{probability of adaptation}. Lastly, we study the time-varying scenario where patches, one after \chg{another}, deteriorate. We consider a third quantity, the \textit{probability of evolutionary rescue}, which corresponds to the probability that a mutant appears by mutation and establishes in the new environment \chg{after} all patches have deteriorated. The theoretical results are complemented by stochastic simulations that support our predictions and help visualize the differences between the different dispersal schemes.

@@

We derive the probability of establishment of a mutant population starting with a single individual initially located either in an old- or a new-habitat patch. In this analysis, we ignore further mutations and are only concerned with the fate of this single mutant lineage. 
The dynamics of the mutant population can be described by a two-type branching process, i.e. \linelabel{R1-27}\chg{all mutants, the descendants of the initial mutant, reproduce, disperse and die independently of each other.} This is a reasonable assumption as long as the overall number of mutants is a lot smaller than the population size of the wild type. The two ``types'' of the two-type branching process correspond to the two habitat types. The process tracks the \linelabel{R2-25}\chg{total number of mutants in old- and new-habitat patches at the end of the life cycle.} The number of offspring of a single mutant can be approximated by Poisson distributed numbers (see eq.~\eqref{eq:sold}). The mean number of offspring of a single mutant, either in an old- or a new-habitat patch, is then given by the following mean reproduction matrix:
%
\begin{equation}\label{eq:mean_repro}
	\bordermatrix{ ~ & \text{old patch} & \text{new patch} \cr
		\text{old patch} & \left(1-m_m^{\text{new}}\right) (1+a_{\text{old}}) & m_m^{\text{new}} (1+a_{\text{new}}) \cr
		\text{new patch} & m_m^{\text{old}} (1+a_{\text{old}}) & \left(1-m_m^{\text{old}}\right) (1+a_{\text{new}})
		}\ ,
\end{equation}
%
where the rows denote the parent locations, and the columns the patch type of the offspring. 
For example, the top-left entry reads as the probability for the parent to stay in an old-habitat patch ($1-m_m^{\text{new}}$) times the average number of offspring in these patches, given by the mean of the corresponding Poisson distribution with rate $(1+a_{\text{old}})$, cf. eq.~\eqref{eq:sold} \linelabel{R2-26}\chg{(see Section S1 in SI for the derivation of $a_{\text{old}}$ at the wild-type equilibrium)}. The other entries are obtained analogously.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Evolutionary Rescue %%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Evolutionary rescue}
Finally, we consider a time-inhomogeneous environment where patches deteriorate one after \chg{another} at regular time intervals $\tau$, until all patches have switched to the new habitat. If the wild-type population fails to generate a successful mutant, the population will \chg{inevitably} go extinct. The probability of evolutionary rescue is therefore tightly linked to the probabilities of adaptation and establishment that we have computed in eqs.~\eqref{eq:estab_approx} and~\eqref{eq:source_sink}. Typically, in formulas expressing the probability of evolutionary rescue, one splits the contributions into mutations arising \textit{de novo} and evolutionary rescue due to standing genetic variation, i.e. mutations that are present in the population before the environmental change \citep{alexander_2014}. We will discuss the effect of standing genetic variation in our system in the following section. For now, we focus on evolutionary rescue due to de novo mutations. We approximate the probability of evolutionary rescue, denoted by $P_{\text{rescue}}$, as
\begin{equation}\label{eq:evol_rescue}
    \begin{aligned}
		P_{\text{rescue}} &\approx 1-\exp\left(- \theta \sum_{i=0}^{M-2} \left(\underbrace{\varphi_{\text{old}}\left(f_{\text{old}}(i)\right) \sum_{j=i\tau}^{(i+1)\tau -1} N_w^{\text{old}}(j)}_{\text{old habitat contribution}} + \underbrace{\varphi_{\text{new}}(f_{\text{old}}(i)) \sum_{j=i\tau}^{(i+1)\tau -1} N_w^{\text{new}}(j)}_{\text{new habitat contribution}}\right)\right.\\
		&\qquad\qquad \qquad\qquad\qquad\qquad\qquad \qquad \qquad \qquad\quad  \left.\underbrace{-\theta \varphi_{\text{new}}(0) \sum_{j = \tau (M-1) }^\infty  N_w^{\text{new}}(j)}_{\text{contribution after the last patch deteriorated}}\right),
	\end{aligned}
\end{equation}
where \linelabel{R1-40}$f_{\text{old}}(i) = (M-i-1)/M$ is the frequency of old-habitat patches after the $(i+1)-$th deterioration event\linelabel{R1-39}\chg{, the establishment probability is given as a function of the patch frequency, $\varphi_k(f_{\text{old}}(i))$,} and $N_w^{k}(j)$ denotes the overall number of wild-type individuals living in habitat $k$ (old or new) in generation~$j$ \chg{(see SI, Section S1 for an approximation)}. The interpretation of this equation is the same as for eq.~\eqref{eq:source_sink}. The only difference is that we now need to account for a changing environment, which alters the population sizes, $N_w^k$, and the establishment probabilities $\varphi_k$ \chg{over time}. In the formula, these changes are accounted for by the sums that iterate through the (discrete) time steps and by the time dependence of the corresponding quantities. We further note that we follow the \chg{expected value of} the wild-type population size deterministically over time, instead of assuming it to be already in its steady state as in eq.~\eqref{eq:source_sink} \chg{(see also Section S1 in the SI)}.

@@

\bibliographystyle{plainnat}
\bibliography{dispersal.bib}

>>>>

\documentclass[11pt]{article}

\usepackage[utf8x]{inputenc}	
\usepackage{fullpage}		
\usepackage[english]{babel}			
%\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fourier}

%\usepackage{ulem}
\usepackage{xcolor}						
\usepackage{amsmath,amstext,amssymb,amsfonts,amsthm,mathrsfs,chemarrow,stackrel,dsfont}   % Math packages
\usepackage{graphicx} 					
\usepackage{subcaption}					
%\usepackage{cite}	
\usepackage[authoryear,sectionbib,sort]{natbib} % Bio citation format
\linespread{1.7}
%\usepackage{titlesec}
%\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
%\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
%\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
%\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}

\usepackage{titlesec}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}

@@

\usepackage[pdftex, breaklinks, colorlinks, citecolor=black, urlcolor=black, linkcolor=black]{hyperref} 
%\usepackage{fourier} % FD I try to avoid the default LaTeX font for non math journals
%\usepackage{FiraSans}
%\usepackage[T1]{fontenc}

@@

%%%APPENDIX
\usepackage[toc,page,header]{appendix}
\usepackage{minitoc}

\usepackage{chapterbib}

%%%
%opening
\title{The effect of habitat choice on evolutionary rescue in subdivided populations}
\author{Peter Czuppon$^{1,2,\ast}$, Fran\c{c}ois Blanquart$^{1,3}$, Hildegard Uecker$^{4,\dag}$,\\ Florence D\'{e}barre$^{2,\dag}$}
\date{}

@@

\doparttoc % Tell to minitoc to generate a toc for the parts
\faketableofcontents % Run a fake tableofcontents command for the partocs

% Make the "Part I" text invisible
\renewcommand \thepart{}
\renewcommand \partname{}

%\part{} % Start the document part
%\parttoc % Insert the document TOC

@@

\linenumbers{}
\modulolinenumbers[2]
\part{}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%% INTRODUCTION %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage

\section*{Introduction}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Evol rescue and dispersal %%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In mathematical models, evolutionary rescue is often studied in a spatially homogeneous situation where the whole population experiences a sudden decrease in habitat quality. In this setting, a large number of theoretical results have been established, for example on the effects of recombination \citep{uecker_2015} and 
horizontal gene transfer \citep{tazzyman_2014}, 
reproduction mechanisms \citep{glemin_2013,uecker_2017}, intra- and interspecific competition \citep{osmond_2013}, predation pressure \linelabel{R1-5}\citep{yamamichi_2015}, 
bottlenecks \citep{martin_2013}, 
different genetic pathways \citep{osmond_2019}, and the context-dependent fitness effects of mutations \citep{anciaux_2018}. \chg{In contrast to these abrupt change scenarios, evolutionary rescue can also be studied in a gradually changing environment \citep[e.g.][]{osmond_2017}}\footnote{\pete{Does anybody know other references?} \hildegard{What about Lynch et al. 1991. Adaptive and  demographic responses of plankton populations to environmental change. Limnology and Oceanography 36:1301--1312., B{\"u}rger and Lynch 1995. Evolution and extinction in a changing environment: a quantitative genetic analysis. Evolution 49:151--163 and Lande and Shannon 1996. The role of  genetic  variation  in adaptation and population persistence in a changing environment. Evolution 50:434--437? But please check; it's a long time that I read those...}}. 
Such gradual changes can in particular occur in fragmented environments.

@@

The population living in this environment consists of asexually reproducing, haploid individuals; generations are discrete and non-overlapping. There are two possible types of individuals, wild types and mutants. The individuals go through the following life-cycle: \begin{enumerate}[label = (\roman*)]
	\item Dispersal: individuals may move between patches. Further details about this step are given below.
	\item Reproduction: individuals reproduce within patches. \linelabel{R2-7}The number of offspring produced by an individual of type $i$ in habitat $k$ (before density regulation, if any), i.e.\ its fecundity, is drawn in a Poisson distribution of expectation $\omega_i^k$. Having fewer than $1$ offspring in expectation means that the local subpopulation will get extinct in the absence of immigration, because the deaths of the parents at each generation are not compensated by enough births on average. On the contrary, the local population is viable if the expected fecundity is greater than $1$. We assume that wild-type and mutant populations are viable in old-habitat patches, and that the mutant's expected fecundity there is lower than the wild type's: $1 \leq \omega^{\text{old}}_m < \omega^{\text{old}}_w$ \footnote{\florence{In the table, it is $0 \leq $, yet the populations are supposed to be viable}}. In new-habitat patches, a wild-type population will eventually go extinct, while a mutant one would persist, hence the term ``rescue mutant'': $\omega^{\text{new}}_w < 1 < \omega^{\text{new}}_m$. 
	All parents die at the end of this step.
	\item \linelabel{R2-9} Mutation: wild-type offspring mutate to the rescue mutant type with probability $\theta$ (back mutations from the mutant to the wild type are neglected).
	\item \linelabel{R2-8}Regulation: if the number of offspring produced locally exceeds the local carrying capacity $K_k$ (where $k$ refers to the habitat type, old or new), the population size is down-regulated to the $K_k$: individuals are randomly removed until the local population size is equal to $K_k$. One mutant offspring has the same chance of being removed as one wild-type offspring: \linelabel{AE-3} we assume that wild-type and mutant individuals are competitively equivalent. If the number of offspring is below the carrying capacity, the regulation step is ignored. We call ``successful offspring'' offspring that survive the density regulation step, and become adults at the next generation. 
	At the end of this step, all offspring become adults, and a new cycle then begins.
\end{enumerate}

@@

\subsection*{Analysis steps}
We decompose our analysis into several steps of increasing complexity. 
\begin{enumerate}
	\item We first consider an environment that is constant over time, and heterogeneous over space, with a fraction $f_{\text{old}}$ of old-habitat patches and $1-f_{\text{old}}$ of new-habitat patches. The population is initiated with wild-type individuals at carrying capacity in old-habitat patches, and at the migration-selection equilibrium $\widehat{N}_w^{\text{new}}$ in new-habitat patches (see Section S1 in the SI for details), 
	and with a single mutant individual, either in an old- or in a new-habitat patch. There are no further mutations ($\theta = 0$), and we compute the \textit{probability of establishment} of a mutant lineage. 
	\item We then consider the same environmental setting, but initialize the population with only wild-type individuals. Mutants can appear by mutation during the simulation ($\theta >0$). We compute the \textit{probability of adaptation}, i.e. that, during a fixed time interval, a mutant appears by mutation and then establishes. 
	\item Finally, we consider the full scenario where each patch degrades one after the other, as described above. The environment is spatially and temporally variable. The population is initialized with only old-habitat patches, all at carrying capacity, with wild-type individuals only. We compute the \textit{probability of evolutionary rescue}, i.e. that a mutant appears by mutation and establishes before the population goes extinct. 
\end{enumerate}

@@

Our calculations assume that the subpopulations in old-habitat patches are always at carrying capacity, i.e.\ that there are always enough offspring that are produced to at least replace all the parents. We also consider that mutants are rare compared to wild-type individuals in old-habitat patches. Denoting by $\widetilde{N}_{i}^{k}$ the number of type-$i$ individuals in a $k$-habitat patch right after dispersal, then the expected number of successful offspring of mutant individuals in this old-habitat patch (i.e., of offspring that survive density regulation and become adults in the next generation) is 
%
\begin{equation}\label{eq:sold}
K_{\text{old}}\, \frac{\omega^\text{old}_m \widetilde{N}^{\text{old}}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w + \omega^\text{old}_m \widetilde{N}^{\text{old}}_m}
%
\approx K_{\text{old}}\, \frac{\omega^\text{old}_m \widetilde{N}^{\text{old}}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w} 
%
= \left(1 + a_{\text{old}}\right) \widetilde{N}^{\text{old}}_m \, .
\end{equation}
%
The approximation results from the assumption that mutants are rare compared to wild-type individuals in old-habitat patches. Eq.~\eqref{eq:sold} defines the per-capita expected growth rate of mutants in old-habitat patches, $a_{\text{old}}$. For the analysis, we also further assume that the number of successful offspring of mutant individuals is Poisson distributed, with mean $1+ a_{\text{old}}$.

\hl{The sign of this local growth rate depends on the value of $\widetilde{N}^{\text{old}}_w$, the size of the local wild-type population right after dispersal. Mutants are less fecund that wild-type individuals in old-habitat patches ($\omega^\text{old}_m < \omega^\text{old}_w$), yet the number of mutants can increase in old-habitat patches if $\widetilde{N}^{\text{old}}_w$ is low enough. In this case indeed, the population is far from carrying capacity, and the offspring of all types of individuals can make it to the next generation. Competition is fiercer as carrying capacity is nearer. }\footnote{\florence{I think this is more Results than Model}}. \linelabel{R1-14}The population dynamics of the wild type are analyzed explicitly in Section~\ref{sec:app:modeldyn} of the Supplementary Information (SI); formulas for $a_{\text{old}}$ are given in eq.~\ref{Seq:s_old}.

In new-habitat patches, on the contrary, we assume that population size remains much lower than carrying capacity, so that all offspring make it to the next generation (no regulation is necessary after reproduction).\footnote{\florence{I am getting rid of notation $r$}} The expected number of successful offspring of mutant individuals in a new-habitat patch, given that there are $\widetilde{N}^{\text{new}}_m$ mutants in the patch right after dispersal, is 
%
\begin{equation}\label{eq:defanew}
\omega_{m}^{\text{new}} \widetilde{N}^{\text{new}}_m = \left(1+a_{\text{new}} \right) \widetilde{N}^{\text{new}}_m,
\end{equation}
which defines $a_{\text{new}}$, the per-capita growth rate of mutants in new-habitat patches. This assumption constrains the differences in carrying capacities among old- and new-habitat patches under which our analysis remains valid. In particular, $K_{\text{new}}$ cannot be too much lower than $K_{\text{old}}$, because in this case, the flow of migrants from old- to new-habitat patches would substantially increase local population sizes in new-habitat patches.

To summarize, for the analysis part, we assume that the number of successful offspring of mutant individuals are Poisson-distributed, with means $(1+ a_{\text{old}})$ in old-habitat patches, and $1+ a_{\text{new}}$ in new-habitat patches, where both quantities are treated as constant over time\footnote{\florence{please check this statement}}. \linelabel{R1-17}\linelabel{R2-13} These assumptions, made for the sake of mathematical analysis, are relaxed in our stochastic simulations.

@@

% Removed because the different steps are now described above
%We investigate the effect of the different dispersal schemes on the probability of %successful adaptation and 
%evolutionary rescue. We first compute the \textit{establishment probability} of a single mutant individual arising either in an old- or in a new-habitat patch. \hl{We then link this probability to the dynamics of a source-sink system}\footnote{\florence{I don't understand what this means -- link?}}, i.e. a fixed environment with a certain number of old habitats (sources) and new-habitat patches (sinks). We derive an expression for the probability of a mutation to emerge and establish in a given time interval. We call this second quantity the \textit{probability of adaptation}. Lastly, we study the time-varying scenario where patches, one after \chg{another}, deteriorate. We consider a third quantity, the \textit{probability of evolutionary rescue}, which corresponds to the probability that a mutant appears by mutation and establishes in the new environment \chg{after} all patches have deteriorated. The theoretical results are complemented by stochastic simulations that support our predictions and help visualize the differences between the different dispersal schemes.

We proceed step-wise towards the computation of a probability of evolutionary rescue. For each step, we first present a mathematical analysis, for then compare our results to the output of simulations that relax the assumptions made for mathematical purposes.

@@

In this first step, we consider that there is initially a single mutant individual in the population, located either in an old- or a new-habitat patch, and we compute the probability of establishment of the mutant population. In this step, we ignore further mutations and are only concerned with the fate of this single mutant lineage.

\subsubsection*{Mathematical analysis}
The dynamics of the mutant population can be described by a two-type branching process. A key assumption of this framework is that \linelabel{R1-27}all mutants reproduce, disperse and die independently of each other. This is a reasonable assumption as long as the overall number of mutants is a lot smaller than the population size of the wild type. The two ``types'' in the name of method, two-type branching process, correspond to the two habitat types. The process tracks the \linelabel{R2-25}total number of mutants in old- and new-habitat patches at the end of each generation. As mentioned above, the numbers of successful offspring of mutant individuals, i.e., of offspring that survive density regulation and become adults at the next generation, are approximated by Poisson-distributed numbers, with means $(1+ a_{\text{old}})$ in old-habitat patches (eq.~\eqref{eq:sold} and eq.~\eqref{Seq:s_old}), and $1+ a_{\text{new}}$ in new-habitat patches (eq.~\eqref{eq:defanew}). After one generation, the mean number of successful offspring of a single mutant, either in an old- or a new-habitat patch, is then given by the following mean reproduction matrix:
%
\begin{equation}\label{eq:mean_repro}
	\bordermatrix{ ~ & \text{old patch} & \text{new patch} \cr
		\text{old patch} & \left(1-m_m^{\text{new}}\right) (1+a_{\text{old}}) & m_m^{\text{new}} (1+a_{\text{new}}) \cr
		\text{new patch} & m_m^{\text{old}} (1+a_{\text{old}}) & \left(1-m_m^{\text{old}}\right) (1+a_{\text{new}})
		}\ ,
\end{equation}
%
where the rows denote the parent locations, and the columns the patch type of the successful offspring. 
For example, the top-left entry is the probability for the parent to stay in an old-habitat patch ($1-m_m^{\text{new}}$, see eq.\eqref{eq:dispersal}) times the average number of successful offspring in these patches \linelabel{R2-26}(see Section~\ref{sec:app:modeldyn} in SI for the derivation of $a_{\text{old}}$ at the wild-type equilibrium). The other entries are obtained analogously.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Evolutionary Rescue %%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Evolutionary rescue}
Finally, we consider a time-inhomogeneous environment where patches deteriorate one after \chg{another} at regular time intervals $\tau$, until all patches have switched to the new habitat. If the wild-type population fails to generate a successful mutant, the population will \chg{inevitably} go extinct. The probability of evolutionary rescue is therefore tightly linked to the probabilities of adaptation and establishment that we have computed in eqs.~\eqref{eq:estab_approx} and~\eqref{eq:source_sink}. Typically, in formulas expressing the probability of evolutionary rescue, one splits the contributions into mutations arising \textit{de novo} and evolutionary rescue due to standing genetic variation, i.e. mutations that are present in the population before the environmental change \citep{alexander_2014}. We will discuss the effect of standing genetic variation in our system in the following section. For now, we focus on evolutionary rescue due to de novo mutations. We approximate the probability of evolutionary rescue, denoted by $P_{\text{rescue}}$, as
\begin{equation}\label{eq:evol_rescue}
    \begin{aligned}
		P_{\text{rescue}} &\approx 1-\exp\left(- \theta \sum_{i=0}^{M-2} \left(\underbrace{\varphi_{\text{old}}\left(f_{\text{old}}(i)\right) \sum_{j=i\tau}^{(i+1)\tau -1} N_w^{\text{old}}(j)}_{\text{old habitat contribution}} + \underbrace{\varphi_{\text{new}}(f_{\text{old}}(i)) \sum_{j=i\tau}^{(i+1)\tau -1} N_w^{\text{new}}(j)}_{\text{new habitat contribution}}\right)\right.\\
		&\qquad\qquad \qquad\qquad\qquad\qquad\qquad \qquad \qquad \qquad\quad  \left.\underbrace{-\theta \varphi_{\text{new}}(0) \sum_{j = \tau (M-1) }^\infty  N_w^{\text{new}}(j)}_{\text{contribution after the last patch deteriorated}}\right),
	\end{aligned}
\end{equation}
where \linelabel{R1-40}$f_{\text{old}}(i) = (M-i-1)/M$ is the frequency of old-habitat patches after the $(i+1)-$th deterioration event\linelabel{R1-39}\chg{, the establishment probability is given as a function of the patch frequency, $\varphi_k(f_{\text{old}}(i))$,} and $N_w^{k}(j)$ denotes the overall number of wild-type individuals living in habitat $k$ (old or new) in generation~$j$ \chg{(see SI, Section S1 for an approximation)}. The interpretation of this equation is the same as for eq.~\eqref{eq:source_sink}. The only difference is that we now need to account for a changing environment, which alters the population sizes, $N_w^k$, and the establishment probabilities $\varphi_k$ \chg{over time}. In the formula, these changes are accounted for by the sums that iterate through the (discrete) time steps and by the time dependence of the corresponding quantities. We further note that we follow the \chg{expected value of} the wild-type population size deterministically over time, instead of assuming it to be already in its steady state as in eq.~\eqref{eq:source_sink} \chg{(see also Section \ref{sec:app:modeldyn} in the SI)}.

@@

\bibliographystyle{amnatnat.bst}
\bibliography{dispersal.bib}

\clearpage

\pagenumbering{arabic}% resets `page` counter to 1
\renewcommand*{\thepage}{S\arabic{page}}

%\titleformat{\section}[hang]{\Large\bfseries\filcenter}{\thesection}{1em}{}
%%%%%% Putting an S before the equation, section and figure numbering
\renewcommand\theequation{S\arabic{equation}}
\renewcommand\thesection{S\arabic{section}}
\renewcommand\thefigure{S\arabic{figure}}

\resetlinenumber
\renewcommand\thelinenumber{S\arabic{linenumber}}
\linenumbers{}
\modulolinenumbers[2]

\appendix

\addcontentsline{toc}{section}{Appendix} % Add the appendix text to the document TOC
\part{Appendix} % Start the appendix part
\parttoc % Insert the appendix TOC

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Model dynamics %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage

\renewcommand{\theequation}{A\arabic{equation}}
\setcounter{equation}{0}  % reset counter 
% redefine the command that creates the equation number.
\renewcommand{\thetable}{A\arabic{table}}
\setcounter{figure}{0}
\setcounter{table}{0}
